% vim: set spell:

\chapter{Preface}

%In what follows, we give a philosophical ground for the remainder of the
%thesis. We start with imprecise notions, and arrive at precise definitions.
%Although it may seem like an imprecise ground at first, it is really a more
%honest ground than a purely ``precise'' ground. By stating the imprecise
%notions that our precise definitions build upon, we state the fundamental
%system belief which we deal within.

\section{Formal systems}

\begin{notion}

A \textbf{formal system} is a system of symbols and rules governing their
manipulation.

\end{notion}

A formal system is a purely mental construct. To be used to any effect in the
real world, a formal system must be realised by a physical system.

\begin{notion}

A \textbf{physical system} is a system of physical symbols and processes
manipulating them. 

\end{notion}

For instance, this document was typeset by a physical system. The symbols in
question are typographical symbols, and the process of typesetting is the
process of their typographic composition.

Out of considerations for the reader, the formal systems presented in this
document deal in the same typographical symbols, and govern their typographical
composition.

The reader, equipped with a writing instrument is another physical system. The
reader is welcomed to realise the formal systems in question, in attempt to
falsify the correctness of their physical realisations in this document. 

\section{Basic definitions and notation}

We begin by introducing some notation for sets, objects, and judgements,
leading to a definition of a formal system, as regarded in this document. The
matters of notation address the typesetting of this document, and bear no
logical significance otherwise.

\subsection{Sets}

\begin{definition}

A \textbf{set} is a collection of distinct elements.

\end{definition}

\begin{notation}

A typographical symbol is \textbf{distinct} from another, if the reader
can at a glance tell a difference between them.

\end{notation}

If the elements of a set can be typeset by distinct typographical symbols, the
set can be typeset elementarily by denoting the elements in a sequence. 

\begin{notation}

The elements of a set are denoted by distinct typographical symbols, which fit
inside the following box: \framebox{\vbox to 12pt {\vfil \hbox to 48pt {} \vfil
}}.

\end{notation}

\begin{notation}

A set is denoted elementarily by denoting the elements in a sequence,
left-to-right, top-to-bottom, separated by commas($,$), and enclosed in
braces($\set{}$).

\end{notation}

For instance, $\set{a,b,c}$ elementarily denotes the set containing exactly the
elements $a$, $b$, and $c$. This notational convention superimposes an order on
the elements of a set. This order bears no logical significance.

\begin{notational-corollary}

The elements of a set, when denoted elementarily in a different order, do not
denote a distinguished set.

\end{notational-corollary}

This notation is useful as it permits the reader to confirm that the elements
are distinct by considering the elements in sequence, and confirming, for each
symbol, that it does not appear more than once in the sequence. That is,
provided that the set is finite.

\begin{notion}

A set is \textbf{finite} if it can eventually be denoted elementarily.

\end{notion}

This notation does not permit to denote an infinite set, and is fairly
impractical for denoting large finite sets. To that end, the reader will be
presented with a procedure by which the elements of the set could be denoted
elementarily, had we the typographical tools and the (possibly infinite) time
to do so. The reader is then welcomed to elementarily denote the set at their
leisure. When reasoning about sets denoted by such a ``generating procedure'',
we will reason about the procedure rather than elements themselves.

\subsection{Objects}

\begin{definition}

An \textbf{alphabet} is a set.

\end{definition}

\begin{definition}

An \textbf{object} is some elements drawn from an alphabet, put in relation.

\end{definition}

\begin{notation}

An object is denoted as a typographical composition of the elements in
question, which fits within the following box:

\begin{center}
\framebox{\vbox to 48pt {\vfil \hbox to 48pt {} \vfil }}
\end{center}

\end{notation}

For instance, given the alphabet $\set{a,b,c}$, we may consider, among
others, the following objects:

\begin{center}
$abc$
\quad\quad\quad
$\begin{matrix}
  & a &   \\
a & a & a \\
  & a &
\end{matrix}$
\quad\quad\quad
\begin{tikzpicture}
\draw (-0.5,0.0) node {$a$};
\draw (-0.7,0.2) node {$a$};
\draw (-0.3,0.2) node {$a$};
\draw (-0.5,0.4) node {$a$};
\draw (0.0,0.9) node {$b$};
\draw (0.2,0.7) node {$b$};
\draw (-0.2,0.7) node {$b$};
\draw (0.0,0.5) node {$b$};
\draw (0.5,0.0) node {$c$};
\draw (0.7,0.2) node {$c$};
\draw (0.3,0.2) node {$c$};
\draw (0.5,0.4) node {$c$};
\end{tikzpicture}

\end{center}

\begin{definition}

An object is \textbf{empty} if no symbols are drawn from the alphabet.

\end{definition}

\begin{notation}

For alphabets not containing the element $\varepsilon$, we denote the empty
object by $\varepsilon$ out of considerations for readability.

\end{notation}

\subsection{Judgements}

\begin{definition}

A \textbf{variable} is an object which is a placeholder for another object.

\end{definition}

\begin{definition}

An object is \textbf{concrete} if it is not a variable.

\end{definition}

\begin{definition}

A \textbf{judgement} is a number of concrete objects, put in relation.

\end{definition}

\begin{notation}

A judgement is a typographical combination of concrete objects, which fits
within the following box:

\begin{center}
\framebox{\vbox to 48pt {\vfil \hbox to 48pt {} \vfil }}
\end{center}

\end{notation}

Judgements come in various forms. For instance, given the objects \symb{0} and
\symb{1}, we may consider, among others, the following judgements:

\begin{table}[h!]
\centering
\begin{tabular}{|l|l|}
\hline
\textbf{Judgement} & \textbf{Reading} \\
\hline
$\symb{0}\;\text{nat}$ & \symb{0} is a natural number. \\
\hline
$\symb{1}=\symb{0}+\symb{1}$ & \symb{1} is the sum of \symb{0} and \symb{1}.\\
\hline
\end{tabular}
\end{table}

\begin{definition}

A \textbf{judgement form} is a number of objects, put in relation, some of
which may be variables.

\end{definition}

\begin{notation}

A \textbf{judgement form} is denoted as a typographical combination of the
objects in question, which fits within the following box:

\begin{center}
\framebox{\vbox to 48pt {\vfil \hbox to 48pt {} \vfil }}
\end{center}

\end{notation}

For instance, given the variables $n$, $n_1$, and $n_2$, we may consider, among
others, the following judgement forms:

\begin{table}[h!]
\centering
\begin{tabular}{|l|l|}
\hline
\textbf{Judgement form} & \textbf{Reading} \\
\hline
$n\;\text{nat}$ & $n$ is a natural number. \\
\hline
$n=n_1+n_2$ & $n$ is the sum of $n_1$ and $n_2$.\\
\hline
\end{tabular}
\end{table}

\begin{definition}

An \textbf{inference rule} is a collection of judgement forms called its
\textbf{premises} and a judgement form called its \textbf{conlusion}.

\end{definition}

\begin{notation}

An inference rule is denoted by denoting the premises in a sequence, separated
by due whitespace, over a denotation of the conclusion. The premises and
conclusion are separated by a horizontal bar.

\end{notation}

For instance, we may consider an inference rule with the single premise
$n\;\text{nat}$ and the conclusion $\symb{s}(n)\;\text{nat}$. This rule reads:
a successor of a natural number is also a natural number.

$$
\judgement[Nat-Rec]{
  n\;\text{nat}
}{
  \symb{s}(n)\;\text{nat}
}
$$

\begin{definition}

An \textbf{axiom} is an inference rule with no premises.

\end{definition}

For instance, we may consider an axiom with the conclusion
$\symb{z}\;\text{nat}$. This rules reads: zero is a natural number.

$$
\judgement[Nat-Z]{
}{
  \symb{z}\;\text{nat}
}
$$

\begin{definition}

A \textbf{judgement form} is defined by a collection of inference rules.

\end{definition}

For instance, the judgement form $n\;\text{nat}$, which reads: $n$ is a natural
number, to be defined by the rules \ruleref{Nat-Rec} and \ruleref{Nat-Z}.

\pagebreak



The procedure by which a set is elementarily denoted, is a matter of exhibiting
a procedure by which distinguished visual symbols are generated for the
elements of the set.

For instance, the natural numbers are generated from the finite alphabet
$\set{0,1,2,3,4,5,6,7,8,9}$. The first 10 symbols of $\mathbb{N}$, are just the symbols in order. The 11th symbol


\begin{definition}

A formal system $F=\chev{A,J}$ is a finite set of symbols $A$, and a finite set
of judgements $J$.

\end{definition}

% A formal system with a finite number of symbols can simulate a formal system
% with an infinite (but enumerable) number of symbols.

\begin{definition}



\end{definition}

\pagebreak

A set can be denoted elementarily, by denoting all the elements of the set, or
by exhibiting a procedure by which the elements can be denoted.


\pagebreak

\input{preface/formal-systems}

\input{preface/audience}

\input{preface/preliminaries}
