% vim: set spell:

\chapter{Complexity}

\begin{quotation}

\footnotesize\sffamily\itshape

\begin{flushright}

This may be a good point to mention that, although I have so far been tacitly
equating computational difficulty with time and storage requirements, I don't
mean to commit myself to either of these measures. It may turn out that some
measure related to the physical notion of work will lead to the most
satisfactory analysis; or we may ultimately find that no single measure
adequately reflects our intuitive concept of difficulty.

\smallbreak

\upshape

--- ALAN COBHAM, {\itshape Logic, Methodology and Philosophy of Science} (1964)

\end{flushright}

\end{quotation}

\begin{quotation}

\footnotesize\sffamily\itshape

\begin{flushright}

In practice, the length of computer computations must be restricted, otherwise
the cost in time and money would be prohibitive.

\smallbreak

\upshape

--- H. E. ROSE, {\itshape Subrecursion: functions and hierarchies} (1984)

\end{flushright}

\end{quotation}


\section{Time}

\subsection{Polynomial Time}

\begin{itemize}

\item Recursive characterization of polytime functions in \cite{rose-1984},
proving certain claims by \cite{cobham-1965}. Both question the relation to the
Grzegorczyk hierarchy \cite{grzegorczyk-1953}.

\item Leivant's paper - A Foundational Delineation of Computational Feasibility.

\item Bellantoni and Cook paper - A NEW RECURSION-THEORETIC CHARACTERIZATION OF
THE POLYTIME FUNCTIONS

\item Niel Jones paper.

\end{itemize}

\subsection{Subpolynomial Time}

\begin{itemize}

\item Some problems, although computable in polynomial time, are still hard to
compute in practice (ICALP'2014, Amir Abboud).

\item Remind of the definitions of $O$, $\Omega$, etc.

\item For each of the below show that every subsequent class is distinct from
the proceeding, and exhibit some ``complete'' problems for these classes.

\end{itemize}

\begin{description}

\item[$\bigOh{1}$ --- Constant]

\item[$\bigOh{\alpha\p{n}}$ --- Inverse Ackermann]

\item[$\bigOh{\log^*{\p{n}}}$ --- Log star]

\item[$\bigOh{\log{\log{\p{n}}}}$ --- Log-log]

\item[$\bigOh{\log{\p{n}}}$ --- Log]

\item[$\bigOh{\log{\p{n}^{\bigOh{1}}}}$ --- Polylog]

\item[$\bigOh{n^c}$, for $0<c<1$ --- Fractional power]

\item[$\bigOh{n}$ --- Linear time]

\item[$\bigOh{n\log^*{\p{n}}}$ --- n log star]

\item[$\bigOh{n\log{\log{\p{n}}}}$ --- n log-log]

\item[$\bigOh{n\log{\p{n}}}$ --- n log n]

\item[$\bigOh{n^2}$ --- quadratic]

\item[$\bigOh{n^3}$ --- cubic]

\item[$\bigOh{n^O(1)}$ --- polynomial]

\end{description}

\subsection{Space}
