\begin{thebibliography}{9} % 9 if < 10 references, 99 if < 100 references, etc.

\section*{Published in Journals}

\bibitem[Grzegorczyk (1953)]{grzegorczyk-1953}

Andrzej Grzegorczyk. Institut Matematiczny Polskiej Akademii Nauk. \emph{Some
classes of recursive functions}. 1953. Rozprawy Matematyczne: IV. Edited by
Karol Borsuk, et al. Published by Polskie Towarzystwo Matematyczne Warszawa in
Poland. Printed in Poland, by Wroclawska Drukarnia Naukowa.

\backrefprint

\bibitem[Hofmann (2000)]{hoffmann-2000}

Martin Hoffmann. Laboratory for Foundations of Computer Science, University of
Edinburgh. \emph{Programming languages capturing complexity classes}. 2000. ACM
SIGACT News, Vol. 31, No. 1, pp. 31--42. Published by the ACM in New York, USA.

\emph{A concise survey of a range approaches to implcit complexity theory
between 1965 and 2000.}

\backrefprint

\section*{Books}

\bibitem[Cormen et al. (2009)]{cormen-et-al-2009}

Thomas H. Cormen, Charles E. Leiserson, Ronald L. Rivest, and Clifford Stein.
Massachusetts Institute of Technology.  \emph{Introduction to Algorithms}, 3rd
ed. 2009. Published by The MIT Press in Cambridge, Massachusetts, USA. ISBN:
978-0-262-03384-8. Printed in the USA.

\backrefprint

\bibitem[Graham et al. (1998)]{graham-et-al-1998}

Ronald Lewis Graham, Donald Ervin Knuth, Oren Patashnik. \emph{Concrete
Mathematics}. Second Edition. 1998. Published by Addison-Wesley Publishing
Company, Inc. ISBN: 978-0-201-55802-9. 24th printing. 2011. Printed in
Westford, Massachusetts, USA.

\backrefprint

\bibitem[Huth \& Ryan (2004)]{huth-ryan-2004}

Michael Huth and Mark Ryan. Imperial College London and University of
Birmingham, United Kingdom. \emph{Logic in Computer Science: Modelling and
Reasoning about Systems}, Second Edition. 2004. Published by Cambridge
University Press in New York, USA. ISBN: 978-0-521-54310-1. 7th printing. 2011.
Printed in the United Kingdom by University Press, Cambridge.

\backrefprint

\bibitem[Mogensen (2010)]{mogensen-2010}

Torben Ægidius Mogensen. Department of Computer Science, University of
Copenhagen. \emph{Basics of Compiler Design}. Anniversary Edition. 2010.
Published through \texttt{lulu.com}. ISBN: 978-87-993154-0-6.

\backrefprint

\bibitem[Rose (1984)]{rose-1984}

H. E. Rose, School of Mathematics, University of Bristol. \emph{Subrecursion:
functions and hierarchies}. 1984. Oxford Logix Guides: 9. Typeset by Joshua
Associates, Oxford.  Published by Clarendon Press, division of Oxford
University Press, in New York, USA. ISBN 0-19-853189. Printed in Great Britain,
by The Thetford Press Ltd.

\backrefprint

\bibitem[Sipser (2013)]{sipser-2013}

Michael Sipser. \emph{Introduction to the Theory of Computation}. Third
Edition. 2013. Published by Cengage Learning. ISBN: 978-1-133-18779-0. Printed
in the USA.

\backrefprint

\section*{In Proceedings}

\bibitem[Cobham (1965)]{cobham-1965}

Alan Cobham. IBM Research Center, Yorktown Heights, New York, USA. \emph{The
intrinsic computational difficulty of functions}. 1965. In Proceedings of the
1964 International Congress for Logic, Methodology and Philosophy of Science,
pp. 24--30. Edited by Yehoshua Bar-Hillel. Published by North-Holland
Publishing Company in Amsterdam, Holland.  Printed in Israel, by Jerusalem
Academic Press Ltd.

\backrefprint

\bibitem[Sazonov (1987)]{sazonov-1987}

V.Yu. Sazonov. Institute of Mathematics, Novosibirsk, USSR. \emph{Bounded set
theory and polynomial computability}. 1987. In Proceedings of the 1987
International Conference on Fundamentals of Computation Theory. Lecture Notes
in Computer Science, Vol. 278, pp. 391--395. Editied by L. Budach, R.G.
Bukharajev, and O.B. Lupanov. Published by Springer-Verlag in Berlin, Germany.
Printed by Druckhaus Beltz.

\backrefprint

\section*{Other}

\bibitem[Bennett (1962)]{bennett-1962}

J.H. Bennett. Ph.D.Thesis, Princeton University. \emph{On Spectra}. 1962.
Princeton, New Jersey, USA.

\emph{Referenced by \cite{cobham-1965}, not attained.}

\backrefprint

\end{thebibliography}
